---
title: "Data Acquisition and Survey Methods"
author: "Group 10"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
---

# Exercise habits and their impact on students' focus

Group 10 investigates whether students' physical exercise habits are related to their reported level of mental brain fog during lectures. The original proposal referred to exercise frequency and brain fog In the final survey data, the available Group 10 variables are `exercise minutes`, `preferred exercise type` and `brain fog`. Therefore, the analysis below focuses on these measured variables.

## Load data {#load-data}

```{r libraries, warning=FALSE, message = FALSE}
library(tidyverse)
library(kableExtra)
```

```{r load data}
# Full dataset
total_survey <- read_csv("Survey_results_2026.csv", show_col_types = FALSE)

# Only Group 10 data
survey <- total_survey %>%
  select(
    starts_with("DQ"),
    starts_with("G10")
  )

# Remove strings
names(survey) <- names(survey) %>%
  str_remove("^DQ\\s*") %>%
  str_remove("^G10\\s*")

# Rename columns
survey <- survey %>%
  rename(
    age = 1,
    gender = 2,
    academic_program = 3,
    exercise_minutes = 4,
    exercise_type = 5,
    brain_fog = 6
  )

# Factorize and create new variables
survey <- survey %>%
  mutate(
    age = factor(age, levels = c("18-20", "21-23", "23-25", "25-30", "30-35", "over 35"), ordered = TRUE),
    exercise_type = as.factor(exercise_type),
    gender = as.factor(gender),
    academic_program = as.factor(academic_program),
    intensity_level = case_when(
      exercise_type %in% c("Weightlifting", "Cardio", "Team Sports") ~ "High-intensity",
      exercise_type == "Yoga/Flexibility" ~ "Low-intensity",
      TRUE ~ NA_character_
    ),
    intensity_level = factor(intensity_level, levels = c("Low-intensity", "High-intensity"))
  )

```

## Survey questions and variables {#survey-questions-and-variables}

The analysis is based on the mandatory demographic questions and the three Group 10 survey questions.

| Variable | Survey question | Type |
|------------------------|------------------------|------------------------|
| `age` | How old are you? | Ordinal categorical |
| `gender` | What is your gender? | Categorical |
| `academic_program` | What is your academic program? | Categorical |
| `exercise_minutes` | How many minutes of physical exercise did you complete in the last 7 days? | Quantitative |
| `exercise_type` | What is your preferred type of exercise? | Categorical |
| `brain_fog` | Rate your average level of mental brain fog during lectures on a scale of 1 = none to 5 = severe. | Ordinal / quantitative scale |

## Research questions and hypotheses {#research-questions-and-hypotheses}

**RQ1:** Do students who complete more minutes of physical exercise report lower levels of brain fog during lectures?\
**Hypothesis:** Exercise minutes and brain fog are negatively associated.

**RQ2:** Does high intensity training (e.g. weightlifting) relate to lower brain fog compared to low intensity training (e.g. yoga)?\
**Hypothesis:** Students preferring high intensity training report lower brain fog than those preferring low intensity training.

**RQ3:** Is age negatively associated with the number of minutes spent on physical exercise per week?\
**Hypothesis:** Older students report fewer exercise minutes.

```{r analysis-objects}
rq1_data <- survey %>%
  filter(!is.na(exercise_minutes), !is.na(brain_fog))

rq2_data <- survey %>%
  filter(!is.na(intensity_level), !is.na(brain_fog))

rq3_data <- survey %>%
  filter(!is.na(age), !is.na(exercise_minutes)) %>%
  mutate(age_numeric = as.numeric(age))

rq1_test <- cor.test(
  rq1_data$exercise_minutes,
  rq1_data$brain_fog,
  method = "spearman",
  alternative = "less",
  exact = FALSE
)

rq2_test <- wilcox.test(brain_fog ~ intensity_level, data = rq2_data, exact = FALSE)

rq3_test <- cor.test(
  rq3_data$age_numeric,
  rq3_data$exercise_minutes,
  method = "spearman",
  alternative = "less",
  exact = FALSE
)

rq3_test_two <- cor.test(
  rq3_data$age_numeric,
  rq3_data$exercise_minutes,
  method = "spearman",
  alternative = "two.sided",
  exact = FALSE
)

p_text <- function(p) {
  ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
}

brain_fog_cols <- c(
  "1" = "#1a9850",
  "2" = "#91cf60",
  "3" = "#FFDB58",
  "4" = "#fc8d59",
  "5" = "#d73027"
)

rq2_summary <- rq2_data %>%
  group_by(intensity_level) %>%
  summarise(
    n = n(),
    mean_brain_fog = mean(brain_fog, na.rm = TRUE),
    median_brain_fog = median(brain_fog, na.rm = TRUE),
    .groups = "drop"
  )

lowest_fog_type <- rq2_summary$intensity_level[which.min(rq2_summary$mean_brain_fog)]
highest_fog_type <- rq2_summary$intensity_level[which.max(rq2_summary$mean_brain_fog)]
```

## Exploratory data analysis {#exploratory-data-analysis}

The exploratory analysis uses one visualization for each research question. The plots include both quantitative and categorical variables and help identify patterns before formal hypothesis testing.

### RQ1: Exercise minutes and brain fog

We built a data visualization to analyze how weekly exercise minutes vary across different levels of brain fog. We converted the `brain_fog` variable into a discrete factor so R would treat it as a categorical x-axis with five distinct groups.

```{r rq1-plot, fig.width=7, fig.height=5}

ggplot(survey, aes(x = factor(brain_fog), y = exercise_minutes, fill = factor(brain_fog))) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  geom_jitter(aes(color = factor(brain_fog)), width = 0.2, alpha = 1, size = 1.5) +
  scale_y_continuous(breaks = c(0,120,240,360,480,600,720,840,960)) +
  scale_fill_manual(values = brain_fog_cols, name = "Brain fog level") +
  scale_color_manual(values = brain_fog_cols, guide = "none") +
  labs(
    title = "Exercise minutes by brain fog level",
    subtitle = "Brain fog level: 1 (None) to 5 (Severe)",
    x = "Brain fog level",
    y = "Minutes of exercise in the last 7 days"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Finding:**

Our initial hypothesis stated that exercise minutes and brain fog are negatively associated, meaning that higher levels of physical activity would correspond to lower levels of cognitive fatigue.

The empirical data, however, reveals a more nuanced, non linear relationship rather than a strict, steady downward trend. The central tendencies, indicated by the median lines, actually peak at a mild brain fog level of 2 (around 200 minutes) and show only a gradual decline through levels 3 and 4, while students reporting zero brain fog at level 1 display a surprisingly low median of approximately 100 minutes. Furthermore, the high variability and vertical spread within levels 1 through 3 including several highly active outliers in level 3 exercising up to 1,000 minutes demonstrate that moderate brain fog is common across a wide range of activity levels.

Nevertheless, our hypothesis holds true at the ultimate extreme of the spectrum. The distribution for level 5 is highly compressed and strictly bounded, showing that students experiencing severe brain fog almost never exceed 200 minutes of weekly exercise, while those who exercise intensely (beyond 400 minutes) rarely report severe cognitive symptoms.

The hypothesis fails as a linear rule because high exercise does not automatically guarantee a clear mind.

### RQ2: Brain fog by exercise intensity

A boxplot is appropriate because the explanatory variable is categorical and the outcome variable is measured on a 1-5 scale. We grouped the exercise types into two intensity levels: - High-intensity (Cardio, Weightlifting, Team Sports) and - Low-intensity (Yoga/Flexibility).

```{r rq2-plot, fig.width=7, fig.height=5}
ggplot(survey[!is.na(survey$intensity_level), ], aes(
  x = intensity_level,
  y = brain_fog,
  color = factor(brain_fog)
)) +
  geom_boxplot(color = "black", outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 1.8) +
  coord_flip() +
  scale_y_continuous(breaks = 1:5) +
  scale_color_manual(
    values = brain_fog_cols,
    name = "Brain fog"
  ) +
  labs(
    title = "Brain fog by exercise intensity",
    subtitle = "Brain fog level: 1 (None) to 5 (Severe)",
    x = "Exercise intensity level",
    y = ""
  ) +
  theme_minimal()
```

**Finding:** The visualization reveals that the median brain fog level is identical (level 3) for both high-intensity and low-intensity exercise groups. However, the high-intensity group has a tighter and lower interquartile range (spanning levels 2 to 3) compared to the low-intensity group (which spans levels 2 to 4). This indicates that while the medians are the same, high-intensity exercise is descriptively associated with a slightly lower overall distribution of brain fog.

### RQ3: Age and exercise minutes

A boxplot is appropriate here because age is an ordered categorical variable (six age bands) while exercise minutes is quantitative. 
A line connecting the group medians makes the direction of the age trend visible before formal testing.

```{r rq3-plot, fig.width=7, fig.height=5}
ggplot(rq3_data, aes(x = age, y = exercise_minutes)) +
  geom_boxplot(fill = "#91cf60", alpha = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.18, alpha = 0.5, size = 1.4, color = "#1a9850") +
  stat_summary(fun = median, geom = "line", aes(group = 1),
               color = "#d73027", linewidth = 0.9) +
  stat_summary(fun = median, geom = "point", color = "#d73027", size = 2.2) +
  labs(
    title = "Exercise minutes by age group",
    subtitle = "Red line connects group medians (older bands to the right)",
    x = "Age group",
    y = "Minutes of exercise in the last 7 days"
  ) +
  theme_minimal()
```

**Finding:** Median weekly exercise peaks in the early-twenties bands (21–23 $\approx$ 190 min, 23–25 $\approx$ 180 min) and then starts going downhill through the older groups. The 18–20 may be very low, but since it only contains 3 entries it shouldn't be taken too seriously. The red median line shows a general downward trend with age, alongside wide spread within every group.Because this is an ordered trend across age bands rather than arbitrary differences between unrelated groups, a Spearman rank correlation was chosen, rather than a group-comparison test such as Kruskal-Wallis.

## Descriptive inference {#descriptive-inference}

The following table summarizes the quantitative variables using sample size, mean, standard deviation, median, minimum, and maximum. The median is useful because exercise minutes may be right-skewed if some students report very high exercise times.
```{r descriptive-numeric-table}
numeric_vars <- survey %>%
  select(exercise_minutes, brain_fog)

numeric_summary <- data.frame(
  Variable = c("Exercise minutes", "Brain fog"),
  Type = "Quantitative",
  Category = "",
  N = as.character(sapply(numeric_vars, function(x) sum(!is.na(x)))),
  Percent = "",
  Mean = as.character(round(sapply(numeric_vars, function(x) mean(x, na.rm = TRUE)), 2)),
  SD = as.character(round(sapply(numeric_vars, function(x) sd(x, na.rm = TRUE)), 2)),
  Median = as.character(round(sapply(numeric_vars, function(x) median(x, na.rm = TRUE)), 2)),
  Min = as.character(round(sapply(numeric_vars, function(x) min(x, na.rm = TRUE)), 2)),
  Max = as.character(round(sapply(numeric_vars, function(x) max(x, na.rm = TRUE)), 2))
)

categorical_table <- function(data, variable, label) {
  data %>%
    filter(!is.na(.data[[variable]])) %>%
    count(.data[[variable]], name = "N") %>%
    mutate(
      Variable = label,
      Type = "Categorical",
      Category = as.character(.data[[variable]]),
      Percent = paste0(round(100 * N / sum(N), 1), "%"),
      Mean = "",
      SD = "",
      Median = "",
      Min = "",
      Max = "",
      N = as.character(N)
    ) %>%
    select(Variable, Type, Category, N, Percent, Mean, SD, Median, Min, Max)
}

categorical_summary <- bind_rows(
  categorical_table(survey, "age", "Age group"),
  categorical_table(survey, "gender", "Gender"),
  categorical_table(survey, "academic_program", "Academic program"),
  categorical_table(survey, "exercise_type", "Preferred exercise type"),
  categorical_table(survey, "intensity_level", "Exercise intensity")
)

summary_table <- bind_rows(
  numeric_summary,
  categorical_summary
)

summary_table %>%
  knitr::kable(
    caption = "Descriptive summary statistics for Group 10 variables",
    booktabs = TRUE,
    align = "lllrrrrrrr"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    full_width = FALSE
  )
```

The next table summarizes the main categorical variable, preferred type of exercise.
```{r descriptive-categorical-table}
exercise_type_summary <- survey %>%
  filter(!is.na(exercise_type)) %>%
  count(exercise_type, name = "n") %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))

exercise_type_summary
```

## Analytic inference {#analytic-inference}

Because brain fog is measured on an ordinal 1-5 scale and exercise minutes may not be normally distributed, Spearman's rank correlation is suitable for RQ1 and RQ3. It tests whether two variables have a monotonic relationship. For RQ2, the Wilcoxon rank-sum test (Mann-Whitney U test) is used because it compares the ordinal brain fog variable between two independent intensity-level groups without assuming normality.

```{r analytic-table}
test_results <- data.frame(
  Research_question = c("RQ1", "RQ2", "RQ3"),
  Test = c(
    "Spearman correlation",
    "Wilcoxon rank-sum test",
    "Spearman correlation"
  ),
  Hypothesis = c(
    "More exercise minutes are associated with lower brain fog",
    "High-intensity exercise correlates with lower brain fog",
    "Older students complete fewer exercise minutes"
  ),
  Estimate = c(
    round(unname(rq1_test$estimate), 3),
    NA,
    round(unname(rq3_test$estimate), 3)
  ),
  Statistic = c(
    round(unname(rq1_test$statistic), 2),
    round(unname(rq2_test$statistic), 2),
    round(unname(rq3_test$statistic), 2)
  ),
  P_value = c(
    p_text(rq1_test$p.value),
    p_text(rq2_test$p.value),
    p_text(rq3_test$p.value)
  ),
  Decision_5_percent = c(
    ifelse(rq1_test$p.value < 0.05, "Reject H0", "Do not reject H0"),
    ifelse(rq2_test$p.value < 0.05, "Reject H0", "Do not reject H0"),
    ifelse(rq3_test$p.value < 0.05, "Reject H0", "Do not reject H0")
  )
)


test_results %>%
  rename(
    RQ = Research_question,
    `P value` = P_value,
    `Decision 5 percent` = Decision_5_percent
    
  ) %>%
  mutate(
    Hypothesis = kableExtra::linebreak(Hypothesis, align = "l")
  ) %>%
  knitr::kable(
    caption = "Analytic inference results",
    booktabs = TRUE,
    escape = FALSE,
    align = "lllrrll"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  ) %>%
  kableExtra::column_spec(1, width = "1cm") %>%
  kableExtra::column_spec(2, width = "3cm") %>%
  kableExtra::column_spec(3, width = "3cm")
```

### Interpretation of analytic results

#### RQ1 result

To address our first research question, we performed a Spearman rank correlation test to evaluate the relationship between weekly physical exercise and reported brain fog levels during lectures. This non-parametric test is given that our independent variable is an ordinal scale from 1 to 5, which violates the normality assumptions required for standard linear models.

The test yielded a Spearman correlation coefficient estimate of $-0.113$, indicating a very weak negative relationship between the two variables. The resulting p-value was calculated to be 0.051. Because this p-value is slightly above our strict 5% significance threshold ($\alpha = 0.05$), we do not reject the null hypothesis ($H_0$), which states that there is no monotonic association between exercise minutes and brain fog.

The statistical test confirms that we cannot definitively claim a general trend of lower brain fog for students who exercise more across the entire dataset. While our visualization hinted at a protective threshold at the extreme ends, specifically that severe brain fog is isolated to lower exercise levels, the weak correlation coefficient and the borderline non significant p-value demonstrate that this pattern is not strong or consistent enough across the central categories to be statistically reliable.

Ultimately, these findings suggest that while high levels of physical activity may limit severe cognitive fatigue for some, minor fluctuations in weekly exercise minutes do not systematically predict lower levels of day to day brain fog for the broader student body.

#### RQ2 result

To evaluate whether high-intensity training correlates with higher brain fog (measured by lower brain fog) compared to low-intensity training, we conducted a Wilcoxon rank-sum test (Mann-Whitney U test). This non-parametric test is appropriate because it compares the distribution of an ordinal variable (brain fog on a 1-5 scale) across two independent groups (High-intensity vs. Low-intensity).

The statistical test yielded a p-value of $0.634$. Since this p-value is well above the standard 5% significance threshold ($\alpha = 0.05$), we fail to reject the null hypothesis. Statistically, there is no significant shift in the distribution of brain fog scores between students who prefer high-intensity workouts and those who prefer low-intensity workouts.

These analytic results align closely with our visual exploratory findings. While the boxplot showed a slightly more concentrated cluster of responses for the high-intensity group, the median brain fog level was identical (level 3) across both intensity brackets. The Wilcoxon test confirms that the observed minor differences in variance are not statistically meaningful.

Ultimately, we must reject the hypothesis that high intensity training provides a superior benefit to brain fog compared to low intensity training.

#### RQ3 result

To test whether older students report fewer weekly exercise minutes, we used a Spearman rank correlation. 
This was chosen because age was collected in ordered bands and not continuous values. Also exercise minutes are right-skewed (mean $\approx$ 195, max = 1100).
Also because the hypothesis is directional, we used a one-sided test.

The test returned a coefficient of $\rho = `r round(unname(rq3_test$estimate), 3)`$, indicating a weak negative monotonic relationship between age and weekly exercise. The one-sided p-value was `r p_text(rq3_test$p.value)`. Because this is below the 5% threshold ($\alpha = 0.05$), we reject the null hypothesis and conclude that the data provide evidence that older students complete fewer minutes of weekly exercise. For transparency, the two-sided p-value is `r p_text(rq3_test_two$p.value)`, i.e. only marginally above 0.05; the one-sided test is justified by the directional hypothesis.

This aligns with the exploratory boxplot, where median minutes drift downward across the older bands and fall most clearly at 30–35. We deliberately use Spearman rather than Kruskal-Wallis because Kruskal-Wallis ignores the ordering of the age bands and only detects group differences, and on this data it is not significant (p $\approx$ 0.28). The contrast shows the effect is a weak monotonic trend with age rather than a large gap between specific groups.

Two things that we should mention are, that the association is weak ($|\rho| \approx 0.14$), so age explains only a small share of the variation in exercise minutes, and the extreme bands are sparse (18–20 has 3 respondents, over-35 has 8). 
Because of this the findings should be treated as an observation and not as a causal effect.

## Functions used {#functions-used}

| Function | Purpose in this report |
|------------------------------------|------------------------------------|
| `read_csv()` | Loads the CSV survey file. |
| `select()` and `starts_with()` | Extract the demographic and Group 10 variables. |
| `str_remove()` | Removes the prefixes `DQ` and `G10` from column names. |
| `rename()` | Gives the selected columns shorter names. |
| `mutate()` | Converts variables and creates analysis-ready values. |
| `filter()` | Removes missing values before plots, summaries, and tests. |
| `ggplot()` | Creates all visualizations. |
| `geom_boxplot()`, `geom_jitter()`, `geom_point()`, `geom_smooth()` | Show distributions, individual observations, scatterplots, and trend lines. |
| `count()` and `summarise()` | Compute categorical frequencies and summary statistics. |
| `cor.test()` | Performs Spearman correlation tests for RQ1 and RQ3. |
| `wilcox.test()` | Wilcoxon rank-sum (Mann-Whitney U) test comparing brain fog between exercise-intensity groups for RQ2. |
| `knitr::kable()` | Displays results in readable tables. |

## Conclusion {#conclusion}

This report explored the relationship between physical exercise habits and mental brain fog among students. The descriptive analysis summarizes how much students exercise, which exercise types they prefer, and how strongly they report brain fog during lectures. The exploratory plots show the observed patterns visually, while the analytic tests evaluate whether these patterns are statistically supported.

Overall, the findings should be interpreted as associations rather than causal effects. The survey is observational, so it cannot prove that exercise directly causes changes in brain fog. However, the results provide useful evidence about whether students with different exercise habits also report different levels of focus-related difficulty during lectures.
